Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-05-04 MIMOSA.tex
Cleaned up the methods section

\section{Introduction}
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). Recently, DVS Sciences Inc. introduced a new cytometry technology named cytometry Time-Of-Flight (CyTOF) that uses rare isotopes instead of fluorochromes (as in standard FCM), as antibody tags to label single cells. CyTOF drastically extends the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize cell poly-functionality. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability. This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood. These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences. Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test)~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}. In the context of single-cell gene expression, very few methods have been proposed. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

@@

\textit{Flow cytometry}: This ICS data is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

\textit{Fluidigm single--cell gene expression}: This is a single--cell gene expression data set of flow--sorted CD8 T--cells from sixteen individuals. The T--cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single--cell level and compared to paired, unstimulated controls.

In the remainder of the paper, we use the following notation to describe our model. 
From this point on, we assume that we observe cell counts from $I$ individuals in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a marker. Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations. We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples. We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individuals $i$ in each sample. For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{sk}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{uk}=\{n_{uik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

\section{Differential expression with one marker}
In practice, stimulation studies like the ones presented here, are often analyzed one marker at a times to avoid being under powered due to the large number of combinations and the potential for very small cell counts in many of the combinations. As a consequence, we first consider the one marker case, \textit{i.e.} $K=1$. In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

@@

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Marker}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}
For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Individuals that show differential expression for a given marker will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described below.

\subsection{Beta--Binomial Model for One Marker}
For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,
\begin{align}
(n_{si}|p_{ui}) \sim \mathrm{Bin}(N_{si},p_{ui});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:null};&\text{if $i$ is a non-responder}\\
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si});\text{ }& (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui});&\text{if $i$ is a responder}\label{eq:alternate}
 \end{align}
 where $p_{si}$, $p_{ui}$ are the unobserved proportions for the stimulated and un-stimulated paired samples. Under the null hypothesis, \textit{i.e.}~\eqref{eq:null}, there is no difference between the stimulation and control samples, and the proportions are equal. Under the alternative model given by \eqref{eq:alternate}, there is a difference in proportions between the two samples.

\subsection{Priors}
Our model shares information across all individuals using an exchangeable Beta priors on the unknown proportions, as follows, 
 \begin{align}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior-null}\\
(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u) \label{eq:prior-alternative}
 \end{align}
where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.  $p_{si}\ne p_{ui}$, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals. Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present for the alternative model. Under the null model, the stimulated condition should behave just like the unstimulated one, and thus the stimulated response rates (\textit{i.e.} $p_{si}$'s) should follow the same prior distribution. Finally, we assume that the $z_i$'s are independent and identically distributed as Bernoulli with probability $w$, where $w$ represents the proportion of responders. It follows that marginally, i.e. after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$. Treating the $z_i$'s as missing data, the unknown parameter vector $\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can easily be estimated via an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}. We will also explore a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ wis assumed to be drawn from a uniform distribution between 0 and 1. In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{sec:estimation}.

\subsection{Parameter estimation}
\label{sec:estimation}
Whether we use an EM or an MCMC algorithm for parameter estimation, we make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p$'s for the null and alternative models, to simplify our calculations. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Appendix A), and are given by,
 \begin{align}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
	&=\prod_{i=1}^P\binom{N_{si}}{n_{si}}\binom{N_{ui}}{n_{ui}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
 \end{align} 
and
\begin{align}
	\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}.\label{model2:unconstrained}
\end{split}
\end{align}

Assuming that the missing data $z_i$'s are known, we define the complete log-likelihood as follows,
\begin{equation}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w)
\label{eq:cll}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.

\subsubsection{EM algoritm}
Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$, and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
\[
\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{\tilde{w}\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+(1-\tilde{w})\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
\]

Replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}, we then optimize the complete-data log-likelihood over $\boldsymbol{\theta}$. Straightforward calculations lead to 
$\tilde w = \sum_i{\tilde{z_i}}/P$, but unfortunately no closed form solutions exist for the remaining parameters. We used numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.  Starting from some initial values, the EM algorithms iterates between the E and M steps until convergence. In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components. We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\subsubsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s that were done via Gibbs samplings.
All full conditionals are given in supplementary material. We used the method of Raftery and Lewis \citep{Raftery1,Raftery2} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

@@

\section{Differential expression across marker combinations}
\subsection{Model}
Our beta-binomial model described in section \ref{sec:} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations. As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{sk}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{uk}=\{n_{uik}: k=1,\dots,2^K\}$. Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.

\begin{table}[h]
\centering
\parbox{0.8\linewidth}{
\caption{Contingency tables for counts of cells expressing two markers between stimulated and unstimulated conditions. $n_{\{s,u\}k}$ denotes observed counts for stimulated or unstimulated table cell $k$, and individual $i$  }\label{tab:multdir}
\begin{minipage}[b]{0.5\linewidth}
\centering
\begin{tabular}{rrr}
\multicolumn{3}{c}{Stimulated}\\
  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Marker A}\\
 & Negative & Positive \\ 
 \multicolumn{1}{l}{Marker B}&&\\
  \hline
Negative &   $n_{si1}$ &   $n_{si2}$ \\ 
Positive &   $n_{si3}$ &   $n_{si4}$ \\ 
   \hline
\end{tabular}
\end{minipage}
\begin{minipage}[b]{0.5\linewidth}
\centering
\begin{tabular}{rrr}
\multicolumn{3}{c}{Unstimulated}\\
  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Cytokine A}\\
 & Negative & Positive \\ 
 \multicolumn{1}{l}{Cytokine B}&&\\
  \hline
Negative &   $n_{ui1}$ &   $n_{ui2}$ \\ 
Positive &   $n_{ui3}$ &   $n_{ui4}$ \\ 
   \hline
\end{tabular}
\end{minipage}
}
\end{table}

In this context, for a given individual,  the beta distribution is replaced by a multinomial distribution, as follows,

>>>>

\section{Introduction}
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize cell poly-functionality. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability. This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single--cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple variables simultaneously on individual cells in a heterogeneous mixture such as whole blood. These variables are used to classify individual cells in the mixture into more homogeneous sub--populations based on phenotypic or functional differences. Such single--cell assays are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

Although there is no standard approach to analyzing ICS assay data current methods range from ad--hoc rules based on log--fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (e.g., Fisher's exact test and $\chi^2$ test\footnote{make sure there is a reference for that})~\cite{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx}. These methods generally test pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}. In the context of single-cell gene expression, very few methods have been proposed. To date, published analysis of Fluidigm data ignore cells where transcripts are undetected, focusing solely on continuous gene expression measurements. However, the proportion of single cells expressing individual genes also carries information and should be evaluated.

@@

\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of flow-sorted CD8 T-cells from sixteen individuals. The T-cells from blocks of four individuals were stimulated with different antigens, gene expression measured at the single-cell level and compared to paired, unstimulated controls. \footnote{Add a bit more info here, how many genes, tetramer sorted, etc}.

In the remainder of the paper, we use the following notation to describe our model. 
From this point on, we assume that we observe cell counts from $I$ individuals in two conditions: stimulated and un-stimulated. Each cell can either be positive or negative for a marker. Given a set of $K$ markers, the measured cells can be classified into $2^K$ positive/negative marker combinations. We denote by $n_{sik}$ and $n_{uik}$, $k=1\dots 2^K$, the observed counts for the $2^K$ combinations in the stimulated and un-stimulated samples. We denote by $N_{si}=\sum_k n_{sik}$ and $N_{ui}=\sum_k n_{uik}$ the total number of cells measured for individual $i$ in each sample. For ease of notations, we will denote by $\mathbf{y}_i$ the vector of observed counts for individual $i$, \textit{i.e.} $\mathbf{y}_{i}=(\mathbf{n}_{si}, \mathbf{n}_{ui})$ where $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$. Finally, we define $\mathbf{y}=(\mathbf{y}_1,\dots,\mathbf{y}_I)$.

\section{Differential expression with one marker}
Datasets like the ones presented here are usually analyzed one marker at a times to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many of the combinations. As a consequence, we first consider the one marker case where cell counts are marginalized and each marker is analyzed separately and $K=1$. In this case, for a given individual, the data can be summarized in a contingency table of $+/-$ cell counts across the un-stimulated and stimulated samples as depicted in Table \ref{tab:twobytwo}.

@@

  \hline
\multicolumn{1}{l}{}&
\multicolumn{2}{c}{Marker}\\
 & Negative & Positive \\ 
  \hline
Stimulated &   $N_{si} - n_{si}$ &   $n_{si}$ \\ 
Unstimulated &   $N_{ui}-n_{ui}$ &   $n_{ui}$ \\ 
   \hline
\end{tabular}
}
\end{table}
For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Individuals that show differential expression for a given marker will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.

\subsection{Beta-Binomial Model for One Marker}
For a given individual $i$, the positive cell counts for the stimulated and un-stimulated samples are jointly modeled as follows,

\begin{equation}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
\begin{equation}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:beta_prior}
\end{equation}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder.

\subsection{Priors}
Our model shares information across all individuals using exchangeable Beta priors on the unknown proportions, as follows, 
 \begin{align}
(p_{ui} | z_{i}=0)  &\sim \mathrm{Beta}(\alpha_u, \beta_u)\label{eq:prior-null}\\
(p_{si} | z_{i}=1)  &\sim \mathrm{Beta}(\alpha_s,\beta_s) \quad\mathrm{and}\quad (p_{ui}|z_{i}=1) \sim \mathrm{Beta}(\alpha_u, \beta_u) \label{eq:prior-alternative}
 \end{align}
where $z_i$ is an indicator variable equal to one if individual $i$ is a responder, i.e.  ${\cal M}_1$ is true, and zero otherwise, and $\alpha_u, \beta_u, \alpha_s,\beta_s$ are unknown hyper-parameters shared across all individuals. Note that the parameters $\alpha_u, \beta_u$ are explicitly shared across the two models, whereas $\alpha_s,\beta_s$ are only present in the alternative model. Finally, we assume that the $z_i$'s are independent and identically distributed Bernoulli with probability $w$, where $w$ represents the proportion of responders. It follows that marginally, \textit{i.e.} after integrating $z_i$, $p_{ui}$ and $p_{si}$ are jointly distributed as a mixture of a one dimensional and a two dimensional Beta distributions with mixing parameter $w$. Treating the $z_i$'s as missing data, the unknown parameter vector $\boldsymbol\theta\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ can be estimated in an Empirical-Bayes fashion using and Expectation-Maximization~\cite{Dempster:1977ul} algorithm as described in Section \ref{sec:estimation}. As an alternative, we will also explore a fully Bayesian model where the hyperparameters $\alpha_u, \beta_u$ and $\alpha_s, \beta_s$ are given vague exponential priors with mean $10^3$, and $w$ is assumed to be drawn from a uniform distribution between 0 and 1. In this case, all parameters will be estimated via a Markov Chain Monte Carlo algorithm as described in Section \ref{sec:estimation}.

\subsection{Parameter estimation}
\label{sec:estimation}
Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, to simplify our calculations. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Supplementary material), and are given by,
 \begin{align}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
	&=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\label{eq:model1post}
 \end{align} 
and
\begin{align}
	\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
=\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}.\label{model2:unconstrained}
\end{split}
\end{align}
Assuming that the missing data, the $z_i$'s, are known, we define the complete log-likelihood as follows,
\begin{equation}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+z_i\log(w)+(1-z_i)\log(1-w)
\label{eq:cll}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated.

\subsubsection{EM algoritm}
Given an estimate of the model parameter vector $\tilde{\boldsymbol{\theta}}=\left\{\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,\tilde{w}\right\}$ and the data $\mathbf{y}$, the E step consists of calculating the posterior probabilities of differential expression, defined by
\[
\tilde z_{i} \equiv \mathrm{Pr}(z_i=1|\mathbf{y},\tilde{\boldsymbol{\theta}})=\frac{\tilde{w} \cdot \mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s,|\mathbf{y}_i)}{(1-\tilde{w})\cdot\mathrm{L}_0(\tilde{\alpha}_u,\tilde{\beta}_u|\mathbf{y}_i)+\tilde{w}\cdot\mathrm{L}_1(\tilde{\alpha}_u,\tilde{\beta}_u,\tilde{\alpha}_s,\tilde{\beta}_s|\mathbf{y}_i)}.
\] 
The M-step then consist of optimizing the complete-data log-likelihood over $\boldsymbol{\theta}$ after replacing $z_i$ by $\tilde{z}_{i}$ in \eqref{eq:cll}. Straightforward calculations lead to 
$\tilde w = \sum_i{\tilde{z_i}}/I$, but unfortunately no closed form solutions exist for the remaining parameters. We use numerical optimization as implemented in R's \textit{optim} function to estimate the remaining parameters.  Starting from some initial values, the EM algorithms iterates between the E and M steps until convergence. In our case, we initialize the $z_{i}$'s using Fisher's exact test to assign each observation to either the null or alternative model components. We then use the estimated $z_i$'s to estimate the $p_{ui}$'s and $p_{si}$'s and use these to set the hyper-parameters to their method-of-moments estimates.

\subsubsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms \citep{Gelfand}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Supplementary material. We used the method of Raftery and Lewis \citep{Raftery1,Raftery2} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

@@

\section{Differential expression across marker combinations}
Our beta-binomial model described in section \ref{sec:DE} can be generalized to a Dirichlet-multinomial model to assess differential expression across multiple marker combinations. As described in the data section, we now have counts for each marker combination, denoted by  $\mathbf{n}_{si}=\{n_{sik}: k=1,\dots,2^K\}$ and $\mathbf{n}_{ui}=\{n_{uik}: k=1,\dots,2^K\}$. 
%Table~\ref{tab:multdir} gives an illustration of the counts with two putative markers A and B.
\subsection{Model}

In our multivariate model, the beta distribution is replaced by a multinomial distribution, as follows,
\begin{equation}
 (\mathbf{n}_{ui}|\mathbf{p}_{ui},) \sim \mathcal{M}(N_{ui},\mathbf{p}_{ui})\quad\text{and}\quad (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(N_{si},\mathbf{p}_{si})\label{eq:mult_likeliehood}
 \end{equation}
%; \text{if $i$ is a non--responder}\\
% &(\mathbf{n}_{ui}|\mathbf{p}_{ui}) \sim \mathcal{M}(\mathbf{p}_{ui},N_{ui}); (\mathbf{n}_{si}|\mathbf{p}_{si}) \sim \mathcal{M}(\mathbf{p}_{si},N_{si});\text{if $i$ is a responder}
%\end{align} 
where $N_{\{s,u\}i}=\sum\limits_{k=1}\limits^{2^K} n_{\{s,u\}ik}$ are the number of cells collected and $\mathbf{p}_{ui}$ and $\mathbf{p}_{si}$ are the unknown proportions for the un-stimulated and stimulated samples.

\subsection{Prior}
As in the one-marker case, we share information across subjects using an exchangeable prior on the unknown proportions. This time the beta priors are replaced by Dirichlet priors, as follows,
\begin{align}
(\mathbf{p}_{ui}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u)\\\nonumber
(\mathbf{p}_{ui}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}_u) \quad \text{and}\quad (\mathbf{p}_{si}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}_s)\label{eq:dir_prior}
\end{align}
where the indicator variable $z_i$ is as defined in \eqref{eq:beta_prior}, i.e. $z_i\sim\mathrm{Be}(w)$ where $w$ is the proportion of responders. As in the beta-binomial case both an EM and MCMC algorithms can be used for parameter estimation. When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}_{\{u,s\}}$ and $w$ as for the beta-binomial model.

\subsection{Parameter estimation}
Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (See supplementary material). For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
\begin{align}
\mathrm{L}_0(\boldsymbol{\alpha}_u|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{ \mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}
\end{align}
where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by 
\begin{align}
\mathrm{L}_1(\boldsymbol{\alpha}_u,\boldsymbol{\alpha}_s|\mathbf{n}_s,\mathbf{n}_u) &= \prod_{i=0}^I\frac{\mathrm{B}(\boldsymbol{\alpha}_{u}+\mathbf{n}_{ui}) \mathrm{B}(\boldsymbol{\alpha}_{s}+\mathbf{n}_{si})}{\mathrm{B}(\boldsymbol{\alpha}_s)\mathrm{B}(\boldsymbol{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_k n_{sik}!} \cdot \frac{N_{ui}!}{\prod_k n_{uik}!}.\label{eq:postmult}
\end{align}
The estimation procedures (both EM and MCMC based) for the multinonial-Dirichlet are the same as for the beta-binomial model except that the number of parameters to be estimated is larger. In our experience, the performance of the EM algorithm greatly deteriorates when $K$ becomes larger than 2. The EM algorithms becomes very dependent on the initial values, and can even fail to converge when good initial values are provided. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and can be used even when $K$ is large. More details about both algorithms are given in Supplementary material.
